\begingroup
\small
\renewcommand{\arraystretch}{1.16}
\setlength{\tabcolsep}{6pt}
\begin{longtable}{@{}>{\raggedright\arraybackslash\bfseries}p{0.30\linewidth}>{\raggedright\arraybackslash}p{0.64\linewidth}@{}}
\caption[Experimental hyperparameters and environment]{Experimental hyperparameters and local running environment used for the analyzed result matrix.}
\label{tab:experiment-hyperparameters}\\
\toprule
\normalfont Item & \textbf{Value used in the experiments} \\
\midrule
\endfirsthead
\toprule
\normalfont Item & \textbf{Value used in the experiments} \\
\midrule
\endhead
Compared setups & Main comparison: Final-HV reward, Hybrid reward, and MPaGE-orig. Component ablations: No budget annealing and No Page-Hinkley, each compared against Final-HV reward. \\
\addlinespace[0.25em]
Benchmark tasks & Bi-TSP, Tri-TSP, Bi-CVRP, and Bi-KP. \\
\addlinespace[0.25em]
LLM-call budgets & $B \in \{25,50,100,200\}$. \\
\addlinespace[0.25em]
Random seeds & 2025, 2026, 2027, 2028, and 2029. \\
\addlinespace[0.25em]
Managed population size &
Bi-TSP, Bi-CVRP, and Bi-KP: 4, 6, 8, and 10 at budgets 25, 50, 100, and 200.\newline
Tri-TSP: 6, 8, 10, and 12 at the same budgets. \\
\addlinespace[0.25em]
Generation target per BMAB generation & The target number of generated offspring is equal to the managed population size. \\
\addlinespace[0.25em]
Initialization limit & The initializer targets one successful population of size \code{pop\_size}; the maximum initialization calls are the larger of 25 and five times \code{pop\_size}. \\
\addlinespace[0.25em]
Variation operators & Two operators are used: mutation and crossover. The operator-level bandit selects between these two operators. \\
\addlinespace[0.25em]
Cluster-level arms & Within a generation, arms are semantic cluster--operator pairs. If the clusterer returns $K$ clusters, the cluster bandit considers up to $2K$ cluster--operator arms. \\
\addlinespace[0.25em]
Selection count & Two parent heuristics are selected when crossover is used. \\
\addlinespace[0.25em]
Budget accounting & Initialization, clustering, generation, and review calls each have unit cost. Review is disabled in the analyzed runs. \\
\addlinespace[0.25em]
Bandit exploration constants & Operator exploration constant $c_{\mathrm{op}}=1.0$ and cluster exploration constant $c_{\mathrm{cluster}}=1.0$. \\
\addlinespace[0.25em]
Budget annealing & Enabled for Final-HV reward and Hybrid reward, with $\gamma_{\mathrm{budget}}=0.5$. The No budget annealing ablation disables this exploration-decay factor. \\
\addlinespace[0.25em]
Reward weights & Quality, diversity, and rank weights are $w_q=1.0$, $w_d=0.3$, and $w_r=0.2$, respectively. \\
\addlinespace[0.25em]
Page--Hinkley settings & Drift threshold $\lambda=0.5$ and slack $\delta=0.005$. The No Page-Hinkley ablation sets the threshold high enough that drift resets are effectively disabled in the analyzed runs. \\
\addlinespace[0.25em]
Generator LLM & \code{gpt-4o-mini} through an external OpenAI-compatible chat-completion API. \\
\addlinespace[0.25em]
Cluster LLM & \code{gpt-4o-mini} through the structured clustering API path. \\
\addlinespace[0.25em]
Generator decoding configuration & Maximum completion length 8,192 tokens, temperature 0.7, streaming disabled, timeout 30 seconds. \\
\addlinespace[0.25em]
Cluster decoding configuration & Structured parse call with a 30-second timeout. \\
\addlinespace[0.25em]
Local running environment & Apple M4 chip with 10-core CPU, 8-core GPU, and 16 GB RAM. The local machine runs orchestration and benchmark evaluation; LLM inference is served by the external API. \\
\bottomrule
\end{longtable}
\endgroup
