\begin{longtable}{
  >{\raggedright\arraybackslash}p{0.24\linewidth}
  >{\raggedright\arraybackslash}p{0.14\linewidth}
  >{\raggedright\arraybackslash}p{0.52\linewidth}
}
\caption{Inventory of generative AI and machine learning techniques in the project.}
\label{tab:technique-inventory}\\
\toprule
Technique & Status & Evidence in the project \\
\midrule
\endfirsthead
\toprule
Technique & Status & Evidence in the project \\
\midrule
\endhead
Large language models & Present & The generator and semantic clusterer are external LLM calls controlled by the BMAB workflow. \\
Prompt engineering & Present & The project uses initialization, mutation, crossover, and semantic-clustering prompt families inherited from MPaGE. \\
Bandits / lightweight reinforcement learning & Present & The scheduler implements UCB-style operator selection, budget-aware cluster selection, and Page--Hinkley drift handling. \\
Execution-based evaluation of LLM-generated code & Present & Generated Python heuristics are executed by task evaluators, and failures can produce null scores. \\
RAG & Not found & No information was found in the project source code or documentation. \\
Fine-tuning, PEFT, LoRA & Not found & No information was found in the project source code or documentation. \\
Diffusion models, GANs & Not found & No information was found in the project source code or documentation. \\
Knowledge distillation & Not found & No information was found in the project source code or documentation. \\
Multi-agent system & Not found & The system contains an LLM--evolution loop and two bandit controllers, but no evidence was found for multiple autonomous agents. \\
\bottomrule
\end{longtable}
