\chapter{Proposed Method}

\providecommand{\methodcode}[1]{\code{#1}}

\section{Research Motivation and Design Rationale}

The proposed method, denoted \bmab{}, is designed as a budget-aware extension of the original \mpage{} framework. The central observation behind this thesis is that LLM-driven heuristic design is not only an optimization problem over generated programs, but also a resource-allocation problem. Each heuristic-generation, clustering, or optional reflection step consumes an LLM call, and therefore consumes time, money, and limited experimental budget. A method that produces high-quality heuristics only after many unproductive calls is less useful than a method that converts early calls into a strong heuristic Pareto front.

The original \mpage{} paper formulates automatic heuristic design for multi-objective combinatorial optimization problems (MOCOPs) as a Language Multi-Criteria Heuristic Design problem. It combines three important ingredients: a SEMO-style heuristic population, Pareto-Front-Grid (PFG) elite selection, and LLM-based semantic clustering of heuristic code \cite{ha2026mpage,laumanns2004semo}. This design is well motivated because it searches for a population of heuristics that trade off solution quality and runtime, rather than a single best heuristic. It also recognizes that code-level diversity is not equivalent to semantic diversity: two heuristics may be syntactically different but implement nearly identical search logic.

However, the baseline \mpage{} workflow allocates generation effort in a mostly predetermined manner. Each generation repeatedly selects elites with PFG, groups promising heuristics into semantic clusters, applies mutation or crossover prompts, evaluates the generated program, and updates the population. This loop is effective for maintaining semantic and objective-space diversity, but it does not explicitly learn which operator or which semantic region is producing the most useful heuristics under the current budget. The stopping condition is also naturally expressed in terms of iterations or sample counts, whereas practical LLM-based search is constrained by a finite number of API calls. Under a strict budget $B$, the system should therefore learn where to spend the remaining LLM calls instead of treating all operators and clusters as equally promising.

\bmab{} therefore keeps the productive heuristic-generation mechanisms of \mpage{} and replaces only the scheduling layer. In particular, the implementation preserves the original task templates, prompt taxonomy, secure evaluator, PFG-based population management, and LLM semantic clustering implementation. The new contribution is a two-level budgeted multi-armed bandit controller that decides which variation operator and which semantic cluster should receive each LLM call. This design follows a conservative research principle: the comparison against \mpage{} is made meaningful by reusing the same generation and evaluation substrate, while changing the decision policy that allocates budget.

\section{Problem Formulation}

\subsection{Inner Multi-Objective Combinatorial Optimization Problem}

Let $\mathcal{X}$ be a finite or discrete feasible set for a MOCOP, and let

\begin{equation}
  f(x) = \left(f_1(x), f_2(x), \ldots, f_M(x)\right), \qquad x \in \mathcal{X},
\end{equation}

where all objectives are minimized. For two feasible solutions $x_a,x_b \in \mathcal{X}$, $x_a$ Pareto-dominates $x_b$, written $x_a \prec x_b$, if

\begin{equation}
  f_i(x_a) \le f_i(x_b) \quad \forall i \in \{1,\ldots,M\},
  \qquad
  \exists j: f_j(x_a) < f_j(x_b).
\end{equation}

The Pareto front is the image of the non-dominated solution set in objective space. Its quality is commonly summarized by the hypervolume indicator $HV_r(\cdot)$ with respect to a reference point $r$ \cite{zitzler1999multiobjective,fonseca2006hv}. For a finite objective set $Y$ in a minimization problem, the hypervolume is the Lebesgue measure of the region dominated by the non-dominated points in $Y$ and bounded by $r$:

\[
  HV_r(Y)
  =
  \lambda\!\left(
    \bigcup_{y \in \mathrm{ND}(Y)}
    [y_1,r_1]\times \cdots \times [y_M,r_M]
  \right),
\]

where $\mathrm{ND}(Y)$ denotes the non-dominated subset and $\lambda(\cdot)$ denotes $M$-dimensional volume. A larger hypervolume means that the obtained Pareto approximation is both closer to the ideal region and more broadly spread across trade-offs. The benchmark evaluators in this thesis use this quantity as the inner solution-quality signal and execution time as the efficiency signal.

\subsection{Outer Heuristic-Design Problem}

Following the LMHD formulation in the original \mpage{} paper, a heuristic $h \in \mathcal{H}$ is treated as a program that maps the current SEMO archive to a new candidate solution:

\begin{equation}
  h : 2^{\mathcal{X}} \rightarrow \mathcal{X}, \qquad x' = h(A),
\end{equation}

where $A$ is the current archive of non-dominated solutions maintained by the inner solver. The evaluator assigns each generated heuristic a vector-valued score

\begin{equation}
  E(h) = \left(e_1(h), e_2(h)\right)
       = \left(-HV_{\text{inner}}(h), T(h)\right).
  \label{eq:heuristic-score}
\end{equation}

Equation~\ref{eq:heuristic-score} should be read as the interface between heuristic design and the inner solver. The LLM does not directly output a TSP tour, CVRP route, or knapsack solution. Instead, it outputs a heuristic program $h$. During evaluation, the inner SEMO-style solver repeatedly calls $h$ with its current archive $A$ and expects one candidate solution $x'$. The quality of $h$ is therefore determined indirectly by the Pareto archive that the inner solver obtains after running with that heuristic.

Here $HV_{\text{inner}}(h)$ is the hypervolume achieved by the solutions produced when $h$ is used inside the benchmark solver, and $T(h)$ is the measured runtime. More concretely, if the inner run using $h$ returns a non-dominated solution archive $A_h$, then $HV_{\text{inner}}(h)$ is computed from the objective vectors $\{f(x):x\in A_h\}$ using the task-specific reference point. This is the \emph{inner hypervolume}: it measures how good one generated heuristic is at solving the underlying combinatorial optimization task.

The first coordinate in $E(h)$ is negated because the outer population-management code is written as a minimization procedure. Thus, a heuristic with a larger inner hypervolume receives a smaller first score coordinate. The second coordinate is runtime, so smaller is also better. This score convention is implemented by the benchmark evaluators in the vendored MPaGE task modules.

At the outer level, the algorithm maintains a population $\mathcal{P}_t \subset \mathcal{H}$ of valid heuristics and aims to approximate the non-dominated set in the heuristic score space. Let $F_t = \{E(h): h \in \mathcal{P}_t\}$ be the outer heuristic front at step $t$. Given an LLM-call budget $B$, \bmab{} is optimized for two related criteria:

\begin{align}
  \max \; HV_{\text{outer}}(B)
  &= \max \; HV_r(F_B), \\
  \max \; \mathrm{AUBC}
  &= \max \; \frac{1}{B}\int_0^B HV_{\text{outer}}(b)\,db,
  \label{eq:aubc-objective}
\end{align}

subject to

\begin{equation}
  \sum_{t=1}^{T} c_t \le B.
\end{equation}

The quantity $HV_{\text{outer}}$ is the \emph{outer hypervolume}: it is computed over heuristic score vectors rather than over task solutions. A high outer hypervolume means that the final population contains useful trade-offs between inner solution quality and runtime, for example heuristics that achieve high inner HV quickly and heuristics that are slower but stronger. The final HV objective measures terminal population quality, while AUBC measures how rapidly the method converts budget into Pareto-front quality over the whole run. In the current experiments, the default cost model is call-based, i.e. one LLM request consumes one budget unit. The budget tracker also supports a token-labeled accounting mode, but the implemented main loop and reported experiments use fixed call costs.

\section{System Overview}

\bmab{} is organized as an adaptive orchestration layer around the MPaGE primitives. Figure~\ref{fig:bmab-architecture-en} summarizes the data and control flow.

\begin{figure}[H]
  \centering
  \resizebox{0.98\linewidth}{!}{%
  \begin{tikzpicture}[
    node distance=1.05cm and 0.82cm,
    box/.style={draw, rounded corners=2pt, align=center, text width=2.9cm, minimum height=0.92cm, font=\small},
    ctrl/.style={draw, rounded corners=2pt, align=center, text width=5.7cm, minimum height=1.05cm, font=\small},
    smallbox/.style={draw, rounded corners=2pt, align=center, text width=2.9cm, minimum height=0.88cm, font=\small},
    data/.style={draw, cylinder, shape border rotate=90, aspect=0.25, align=center, text width=2.7cm, minimum height=0.92cm, font=\small},
    arr/.style={-{Latex[length=2.2mm]}, thick}
  ]
    \node[box] (task) {Task template\\and MPaGE prompts};
    \node[box, right=of task] (llm) {Generator LLM\\heuristic code};
    \node[box, right=of llm] (eval) {Secure evaluator\\score and runtime};
    \node[data, right=of eval] (pop) {Population\\PFG archive};

    \node[smallbox, below=of task] (budget) {BudgetTracker\\hard call budget};
    \node[ctrl, below=of llm, xshift=1.9cm] (controller) {Adaptive BMAB controller\\operator UCB + cluster UCB\\cluster priors + reward + Page--Hinkley};
    \node[smallbox, below=of pop] (profiler) {BMABProfiler\\HV curve and logs};

    \draw[arr] (task) -- (llm);
    \draw[arr] (llm) -- (eval);
    \draw[arr] (eval) -- (pop);
    \draw[arr] (pop) -- (profiler);

    \draw[arr] (budget.north) -- (task.south);
    \draw[arr] (budget.east) -- (controller.west);
    \draw[arr] (eval.south) -- ++(0,-0.35) -| (controller.north);
    \draw[arr] (controller.north west) -- ++(0,0.35) -| (llm.south);
  \end{tikzpicture}%
  }
  \caption{Logical architecture of \bmab{}. The method preserves MPaGE's prompt, evaluation, semantic clustering, and population-management components, while adding budget accounting and adaptive bandit control.}
  \label{fig:bmab-architecture-en}
\end{figure}

The implementation is centered in \methodcode{bmab\_llm.py}. The main supporting modules are summarized in Table~\ref{tab:method-modules}.

\begin{table}[H]
  \centering
  \caption{Main methodological modules in the implementation.}
  \label{tab:method-modules}
  \begin{tabular}{p{0.28\linewidth}p{0.62\linewidth}}
    \toprule
    Module & Methodological role \\
    \midrule
    \methodcode{bmab\_llm.py} & Main orchestrator. It owns the evolutionary loop, calls PFG selection, requests semantic clusters, selects bandit arms, samples offspring, computes rewards, and flushes the final population. \\
    \methodcode{budget.py} & Hard budget tracker. It stores total, remaining, and consumed budget and records a history entry for every charged LLM call. \\
    \methodcode{bandit.py} & Two-level adaptive scheduler: persistent operator UCB, per-generation cluster UCB, and Page--Hinkley drift detection. \\
    \methodcode{cluster\_manager.py} & Wrapper around MPaGE semantic clustering. It charges cluster calls, validates or repairs partitions, and computes cluster-quality priors. \\
    \methodcode{reward.py} & Scalar credit assignment for bandit learning, including HVI, final managed-population HV gain, diversity gain, rank smoothing, and invalid-code penalties. \\
    \methodcode{profiler.py} & Records budget-vs-HV curves, AUBC, budget history, bandit state, and population artifacts for analysis. \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Components Inherited from MPaGE}

\bmab{} intentionally retains the core generative and evaluative substrate of \mpage{}. This inheritance is documented in the implementation materials and is visible in the vendored MPaGE subtree. The retained components are important because they isolate the proposed contribution to budget-aware scheduling rather than changing the benchmark, prompt templates, or evaluation logic.

\subsection{Population Representation and PFG Selection}

Each individual is represented as a Python function together with a natural-language algorithm description, evaluation score, sample time, and execution time. The population class inherited from MPaGE maintains a managed set of heuristics using non-dominated sorting and diversity-preserving truncation.

PFG selection partitions the two-dimensional heuristic objective space into grid cells. Let $P_t = \{E(h_i): h_i \in \mathcal{P}_t\}$ be the current score set. For each objective $j \in \{1,2\}$, define the ideal and nadir values

\begin{equation}
  z_j^{\ast} = \min_{h \in \mathcal{P}_t} e_j(h), \qquad
  z_j^{n} = \max_{h \in \mathcal{P}_t} e_j(h).
\end{equation}

With $K_j$ segments and a small margin $\sigma$, the grid width is approximately

\begin{equation}
  \Delta_j = \frac{z_j^{n} - z_j^{\ast} + \sigma}{K_j}.
\end{equation}

A heuristic is assigned to a grid index

\begin{equation}
  g_j(h)=\left\lfloor \frac{e_j(h)-z_j^{\ast}}{\Delta_j}\right\rfloor,
  \qquad
  G(h)=(g_1(h),g_2(h)).
\end{equation}

Within each occupied cell, non-dominated representatives are retained. The union of retained representatives forms the elite set used for reproduction. This mechanism is inherited from \mpage{} and is implemented in the vendored MPaGE population manager. In the current BMAB loop, the PFG-selected slice is used as a high-quality signal for parent weighting, while the wider managed population is made available to the clustering layer to avoid over-narrow clusters.

\subsection{LLM-Based Semantic Clustering}

The original \mpage{} paper argues that objective-space or syntax-only diversity is insufficient for heuristic design. Two code snippets can have different syntax but the same algorithmic logic, and two heuristics with similar performance may use distinct search strategies. MPaGE addresses this by prompting an LLM to group elite heuristics according to semantic and behavioral similarity.

Formally, for a candidate set $P=\{h_1,\ldots,h_n\}$, the clustering module returns

\begin{equation}
  \mathrm{SemClust}(P)=\{C_1,\ldots,C_K\},
\end{equation}

such that

\begin{equation}
  \bigcup_{k=1}^{K} C_k=P, \qquad C_i \cap C_j = \emptyset \; (i \ne j).
\end{equation}

The implementation calls the MPaGE cluster prompt through \methodcode{cluster\_manager.py}. If the LLM response is malformed, incomplete, or unavailable, the method falls back to singleton clusters. This design preserves semantic clustering when possible while ensuring that the evolutionary loop remains robust to LLM-format failures.

\subsection{Mutation and Crossover Prompt Families}

MPaGE defines a prompt taxonomy for initialization, mutation, and crossover. \bmab{} keeps this taxonomy. Initialization uses the I1 prompt to create zero-shot heuristics. Mutation uses M1 or M2 to modify a single parent from the selected cluster. Crossover uses E1 or E2 to combine one parent from the selected cluster with a second parent outside that cluster. Thus, BMAB preserves MPaGE's core semantic rule:

\begin{equation}
  \text{mutation} = \text{intra-cluster local refinement},
  \qquad
  \text{crossover} = \text{inter-cluster recombination}.
\end{equation}

The difference is not the prompt content but the policy that decides when a mutation or crossover prompt should be used, and which cluster should provide the parent.

\section{Hard Budget Model}

The LLM-call budget is treated as a first-class constraint. Let $B$ be the total budget and $b_t$ be the remaining budget before action $t$. Each LLM action has cost $c_t$, and the transition is

\begin{equation}
  b_{t+1} = b_t - c_t, \qquad b_t \ge c_t.
\end{equation}

If the system cannot afford an action, the action is not issued. The current implementation charges budget before the heuristic-generation call or cluster call is sent. This is important experimentally: invalid code, parsing failure, timeout, and non-improving offspring are still charged because the API resource has already been consumed.

In the call-based mode used by the reported experiments, the following costs are used:

\begin{table}[H]
  \centering
  \caption{Budget costs used by the current implementation.}
  \label{tab:budget-costs}
  \begin{tabular}{lc}
    \toprule
    Action & Cost \\
    \midrule
    Initialization call (I1) & $1$ \\
    Heuristic-generation call (M1/M2/E1/E2) & $1$ \\
    Semantic clustering call & $1$ \\
    Optional review/suggestion call & $1$ \\
    \bottomrule
  \end{tabular}
\end{table}

The budget state is recorded through \methodcode{BudgetTracker.history}, and the profiler later writes \methodcode{budget\_history.json}. This gives the experimental harness a direct account of how the budget was spent.

\section{Two-Level Budgeted Bandit Scheduler}

\subsection{Why a Two-Level Scheduler is Needed}

A direct bandit over individual heuristics is not viable because the heuristic space is open-ended: every LLM call can create a previously unseen program. A direct persistent bandit over semantic cluster IDs is also problematic because clusters are reconstructed at each generation and their numeric labels are not stable across generations. \bmab{} therefore separates the adaptive decision into two levels:

\begin{enumerate}
  \item a persistent operator-level bandit over the stable action set $\mathcal{O}=\{\mu,\chi\}$, where $\mu$ denotes mutation and $\chi$ denotes crossover;
  \item a per-generation cluster-level bandit over arms $(k,o)$, where $k$ indexes one semantic cluster in the current generation and $o \in \mathcal{O}$ is the selected operator.
\end{enumerate}

This decomposition lets the method learn long-term operator utility while still adapting to the short-lived semantic cluster structure of the current population.

\subsection{Operator-Level UCB}

For each operator $o \in \mathcal{O}$, the operator bandit stores the number of pulls $n_o$, cumulative reward $S_o^R$, and cumulative cost $S_o^c$. Its empirical utility is reward per cost:

\begin{equation}
  \hat{\rho}_o = \frac{S_o^R}{\max(S_o^c,\varepsilon)}.
\end{equation}

The selected operator maximizes a UCB score:

\begin{equation}
  U_{\mathrm{op}}(o)
  =
  \hat{\rho}_o
  +
  c_{\mathrm{op}}
  \sqrt{\frac{2\log N_{\mathrm{op}}}{n_o}},
  \label{eq:operator-ucb}
\end{equation}

where $N_{\mathrm{op}}=\sum_{o' \in \mathcal{O}} n_{o'}$ and $c_{\mathrm{op}}$ controls exploration. The implementation initializes each operator with an optimistic prior count and prior reward, preventing division by zero and encouraging early exploration. This is implemented in \methodcode{bandit.py} by the \methodcode{OperatorBandit} class.

\subsection{Cluster-Level Budgeted UCB}

At the beginning of each generation, \methodcode{ClusterManager} returns a partition $\{C_1,\ldots,C_K\}$ and a quality prior $q_k$ for each cluster. At this point, $q_k$ can be understood as a scalar estimate of how promising cluster $C_k$ is before any new offspring are generated in the current generation; its formal front-aware definition is given in Section~\ref{sec:front-aware-priors}. The cluster bandit then creates arms $(k,o)$ for all clusters and operator labels. Each arm stores cumulative reward and cumulative cost, and its exploitation term is

\begin{equation}
  \hat{\rho}_{k,o}
  =
  \frac{S_{k,o}^R}{\max(S_{k,o}^c,\varepsilon)}.
\end{equation}

The cluster-level UCB score is

\begin{equation}
  U_{\mathrm{cl}}(k,o)
  =
  \hat{\rho}_{k,o}
  +
  c_{\mathrm{cl}}
  \alpha_t
  \sqrt{\frac{2\log N_{\mathrm{cl}}}{n_{k,o}}},
  \label{eq:cluster-ucb}
\end{equation}

where

\begin{equation}
  \alpha_t =
  \left(\frac{b_t}{B}\right)^{\gamma_b}
\end{equation}

when budget annealing is enabled, and $\alpha_t=1$ otherwise. The factor $\alpha_t$ decreases as the budget is exhausted, making the controller more exploitative near the end of the run. This directly addresses the finite-horizon nature of the task: late calls have little time to recover from risky exploration.

The cluster bandit is warm-started from both cluster quality and operator priors. Let $p_o$ be the softmax-normalized operator utility from the persistent operator bandit. The initial reward assigned to arm $(k,o)$ is proportional to

\begin{equation}
  r^{(0)}_{k,o} = \frac{1}{2}\left(q_k+p_o\right).
\end{equation}

This warm start transfers information from the persistent operator layer into the generation-local cluster layer without assuming that cluster IDs persist across generations.

\section{Front-Aware Cluster Priors}
\label{sec:front-aware-priors}

The semantic cluster partition identifies behavioral groups, but it does not by itself rank those groups. The current implementation therefore computes a front-aware prior for each cluster. For cluster $C_k$, define $S_k=\{E(h):h \in C_k\}$ and $S=\{E(h):h \in \mathcal{P}_t\}$. With outer reference point $r$, the hypervolume contribution of the cluster is

\begin{equation}
  \Delta HV_k
  =
  \max\left(0,\;HV_r(S)-HV_r(S\setminus S_k)\right).
\end{equation}

The heuristic scores in this section are the two-dimensional evaluator outputs $s=E(h)=(-HV_{\text{inner}}(h),T(h))$. Therefore, $s_1$ represents negated inner hypervolume and $s_2$ represents runtime. Because the first score coordinate is negative inner HV, a simple quality proxy for solution strength is

\begin{equation}
  I_k = \max\left(0,\;-\min_{s \in S_k} s_1\right),
\end{equation}

with a finite cap in the implementation to avoid an excessively large prior. A runtime proxy is defined as

\begin{equation}
  T_k = \frac{1}{1+\min_{s \in S_k} s_2}.
\end{equation}

The unnormalized prior is

\begin{equation}
  \tilde{q}_k = \Delta HV_k + 0.25 I_k + T_k,
\end{equation}

and the normalized prior used by the bandit is

\begin{equation}
  q_k = \frac{\tilde{q}_k}{\max_j \tilde{q}_j}
\end{equation}

when the denominator is positive. This design is implemented in \methodcode{cluster\_manager.py}. It is more aligned with the final outer-front objective than a one-dimensional score proxy because it rewards clusters that currently contribute to the heuristic Pareto front while still accounting for inner solution quality and runtime.

\section{Reward and Credit Assignment}

\subsection{Reward Components}

The reward function converts the outcome of one LLM generation call into a scalar signal for the two bandits. For a valid generated heuristic with score $s$, the reward is

\begin{equation}
  R(s)
  =
  w_q Q(s)
  +
  w_d \max(0,\Delta D)
  +
  w_r \mathrm{Rank}(s).
  \label{eq:reward}
\end{equation}

If the generated heuristic is invalid, cannot be parsed, fails evaluation, or returns a non-finite score, the reward is

\begin{equation}
  R(s) = -\lambda_{\mathrm{pen}}.
\end{equation}

The values used in the current implementation are $w_q=1.0$, $w_d=0.3$, $w_r=0.2$, and $\lambda_{\mathrm{pen}}=1.0$. These settings are exposed through the command-line interface for the weights and fixed in \methodcode{RewardComputer} for the penalty.

\subsection{Quality Signal}

Let $F$ be the set of valid population scores before the new candidate is inserted. The immediate hypervolume improvement is

\begin{equation}
  \mathrm{HVI}(s;F)
  =
  \max\left(0,\;HV_r(F\cup\{s\})-HV_r(F)\right).
\end{equation}

The implementation also computes a managed-population HV delta. Let $\mathrm{Manage}_N(\cdot)$ denote the same score-only survivor-selection logic used by the managed population with cap $N$. Then

\begin{equation}
  \Delta HV_{\mathrm{managed}}
  =
  \max\left(
    0,\;
    HV_r(\mathrm{Manage}_N(F\cup\{s\}))
    -
    HV_r(\mathrm{Manage}_N(F))
  \right).
  \label{eq:managed-hv}
\end{equation}

Both quantities are normalized by rolling maxima over a recent window. Let $\widehat{\mathrm{HVI}}$ and $\widehat{\Delta HV}_{\mathrm{managed}}$ denote the normalized values. The quality mode is then

\begin{equation}
  Q(s)=
  \begin{cases}
    \widehat{\mathrm{HVI}}(s;F), & \mathrm{mode}=\mathrm{dense},\\
    \frac{1}{2}\widehat{\mathrm{HVI}}(s;F)
    +\frac{1}{2}\widehat{\Delta HV}_{\mathrm{managed}}(s;F),
    & \mathrm{mode}=\mathrm{hybrid},\\
    \widehat{\Delta HV}_{\mathrm{managed}}(s;F),
    & \mathrm{mode}=\mathrm{final\_hv}.
  \end{cases}
\end{equation}

The experimental framework evaluates three parallel quality-signal variants while retaining the same BMAB scheduling pipeline. Final-HV reward uses \methodcode{final\_hv}, Dense reward uses \methodcode{dense}, and Hybrid reward uses \methodcode{hybrid}. This design makes it possible to compare whether terminal managed-population HV, immediate HVI, or a mixture of the two provides the most useful credit signal.

\subsection{Diversity and Rank Smoothing}

Pure HVI is often sparse: many valid offspring do not immediately expand the outer Pareto front. To provide a denser signal, \bmab{} adds rank and diversity terms. The cumulative diversity index used in the reward is the mean pairwise Euclidean distance among valid score vectors:

\begin{equation}
  D(F)
  =
  \frac{1}{|F|(|F|-1)}
  \sum_{i \ne j} \|s_i-s_j\|_2.
\end{equation}

The diversity increment is $\Delta D = D(F\cup\{s\})-D(F)$. Only positive increments are rewarded. The rank score is

\begin{equation}
  \mathrm{Rank}(s)
  =
  \frac{1}{|F|}
  \left|\{u \in F : \exists j,\; s_j < u_j\}\right|.
\end{equation}

In this equation, $s$ and $u$ are heuristic scores of the form $E(h)=(-HV_{\text{inner}}(h),T(h))$. The first coordinate measures negated inner hypervolume, and the second coordinate measures runtime. Since both coordinates are minimized in the outer heuristic population, a smaller coordinate value indicates either higher inner solution quality or lower execution time. The rank term therefore rewards a candidate that improves at least one objective relative to existing population members, even when it does not immediately increase HV.

\section{Non-Stationarity and Page--Hinkley Reset}

The reward distribution in evolutionary heuristic design is non-stationary. A cluster that is useful early may become unproductive after its design pattern has already been exploited. To reduce overconfidence in stale arm statistics, each cluster-level arm is monitored by a Page--Hinkley detector \cite{page1954continuous}.

For a reward sequence $(r_t)$ of one arm, the implementation maintains the running mean $\bar{r}_t$ and cumulative statistic

\begin{align}
  \bar{r}_t &= \bar{r}_{t-1} + \frac{r_t-\bar{r}_{t-1}}{t}, \\
  m_t &= m_{t-1}+r_t-\bar{r}_t+\delta, \\
  M_t &= \max(M_{t-1},m_t).
\end{align}

Downward drift is declared when

\begin{equation}
  M_t - m_t > \lambda_{\mathrm{PH}}.
\end{equation}

When drift is detected, the corresponding arm statistics are reset to an optimistic prior for the current generation. This mechanism is implemented in \methodcode{bandit.py}. It is limited to the cluster-level bandit because cluster rewards are local to the current generation and most sensitive to rapid changes in population state.

\section{Parent Selection and Variation Workflow}

The original MPaGE strategy uses intra-cluster mutation and inter-cluster crossover. \bmab{} preserves this logic but changes how parents are sampled. The loop first obtains PFG-selected elites through \methodcode{Population.selection}. It then expands the clustering pool to the entire current managed population to give the cluster bandit a richer set of semantic groups. To prevent the wider pool from diluting the PFG signal, parent sampling gives candidates that also appear in the PFG-selected elite set a larger weight:

\begin{equation}
  w(h)=
  \begin{cases}
    3, & h \in E_{\mathrm{PFG}},\\
    1, & \text{otherwise}.
  \end{cases}
\end{equation}

For mutation, a single parent is sampled from the selected cluster using these weights. For crossover, the first parent is sampled from the selected cluster and the second parent is sampled from other clusters when possible. This preserves semantic recombination while biasing the process toward strong PFG representatives. In the implementation, this logic is used when constructing the mutation or crossover prompt and when sampling parents from the selected semantic groups.

\section{Complete Algorithm}

Algorithm~\ref{lst:bmab-algorithm} gives the high-level workflow implemented by \methodcode{BMABLLM.run}. The pseudocode abstracts away exception handling and API-specific details, but preserves the execution order that matters methodologically: initialization, budget-aware clustering, adaptive operator and cluster selection, reward computation, bandit update, drift handling, and final population flushing.

\begin{lstlisting}[language={},caption={High-level BMAB-LLM workflow.},label={lst:bmab-algorithm}]
Algorithm BMAB-LLM
Input: evaluator E, generator L, clusterer C,
       budget B, population cap N, reference point r
Output: managed heuristic population P

initialize BudgetTracker(B), Population P
initialize OperatorBandit, ClusterBandit, RewardComputer

while initialization target is not reached
      and one generation call is affordable do
    charge one budget unit
    h <- L(I1 initialization prompt)
    score <- E(h)
    if score is valid then
        register h in P
    end if
end while

record the initial budget-HV point

while budget remains do
    E_pfg <- PFG-selected elites from P
    H_all <- current managed population

    if a clustering call and a generation call are affordable then
        charge one budget unit
        clusters <- C(H_all)
        if clusters are invalid then
            clusters <- singleton fallback clusters
        end if
    else
        clusters <- singleton fallback clusters
    end if

    q <- front-aware cluster priors
    p <- operator priors from OperatorBandit
    reset ClusterBandit using clusters, q, and p

    produced <- 0
    while produced < N
          and one generation call is affordable do
        op <- select mutation or crossover
        arm <- select a cluster arm for op
        parents <- weighted parent sampling from clusters
        prompt <- MPaGE mutation or crossover prompt

        F_before <- scores in P and pending valid offspring
        D_before <- diversity(F_before)

        charge one budget unit
        h_child <- L(prompt)
        score <- E(h_child)
        if score is valid then
            register h_child as pending
        end if

        F_after <- scores in P and pending valid offspring
        D_after <- diversity(F_after)
        reward <- RewardComputer(score, F_before,
                                 D_before, D_after)

        update OperatorBandit(op, reward)
        update ClusterBandit(arm, reward)
        apply Page-Hinkley reset if drift is detected
        produced <- produced + 1
    end while

    merge pending valid offspring into P when needed
    record bandit state and budget-HV point
end while

flush all pending valid offspring into P
write budget history, budget curve, bandit log, and AUBC
\end{lstlisting}

\section{Baseline and Variant Design}

The MPaGE-orig baseline is not a new algorithm. It is a budget-instrumented wrapper around the original MPaGE driver that records budget consumption and budget-HV curves using the same profiler infrastructure as \bmab{}. The purpose is to compare allocation policies while keeping the task evaluators and generated artifacts as consistent as possible.

The current CLI defines the following method variants in \methodcode{main.py}. The first three BMAB-style variants share the same scheduler and differ only in the quality signal $Q(s)$ used inside the reward:

\begin{table}[H]
  \centering
  \caption{Main method variants relevant to the proposed framework.}
  \label{tab:method-ablation}
  \small
  \begin{tabular}{>{\raggedright\arraybackslash}p{0.29\linewidth}>{\raggedright\arraybackslash}p{0.61\linewidth}}
    \toprule
    Variant & Interpretation \\
    \midrule
    Final-HV reward & BMAB pipeline with the \methodcode{final\_hv} quality signal, operator bandit, cluster bandit, Page--Hinkley, budget annealing, and PFG-weighted parent sampling. \\
    Dense reward & BMAB pipeline with the immediate-HVI quality signal. \\
    Hybrid reward & BMAB pipeline with a half immediate-HVI and half managed-final-HV quality signal. \\
    \methodcode{no\_budget\_anneal} & BMAB pipeline without remaining-budget scaling in the cluster UCB exploration term. \\
    \methodcode{no\_ph} & Disables practical Page--Hinkley resets by using an extremely large threshold. \\
    \methodcode{no\_diversity} & Removes the CDI diversity term by setting its weight to zero. \\
    \methodcode{op\_only} & Disables the cluster bandit and samples clusters uniformly. \\
    \methodcode{cluster\_only} & Disables the operator bandit and uses a round-robin operator sequence. \\
    MPaGE-orig & Budget-instrumented wrapper around the original MPaGE implementation. \\
    \bottomrule
  \end{tabular}
\end{table}

These variants are important because they separate the contribution of each methodological component. In particular, Final-HV reward, Dense reward, and Hybrid reward vary only the reward quality mode, while \methodcode{no\_budget\_anneal}, \methodcode{no\_ph}, \methodcode{no\_diversity}, \methodcode{op\_only}, and \methodcode{cluster\_only} isolate specific control mechanisms.
