\chapter{Conclusion and Future Work}

\section{Summary}

This thesis analyzed \texttt{mpage\_bmab}, a budget-aware extension of MPaGE for LLM-based heuristic design in multi-objective combinatorial optimization. The project preserves MPaGE's core generative and evaluative mechanisms, including SEMO-style archive updates, Pareto Front Grid selection, semantic clustering, prompt-based heuristic generation, and benchmark evaluation. It extends this foundation with an adaptive scheduling layer that treats LLM calls as a scarce resource.

The proposed \bmab{} method can be summarized as follows. Each generator or clusterer invocation is charged by a hard \texttt{BudgetTracker}. An operator-level UCB controller chooses between mutation and crossover. A cluster-level budgeted UCB controller selects semantic cluster--operator arms within each generation. Page--Hinkley drift detection resets stale cluster statistics when reward behavior changes. A bounded reward function converts heuristic-evaluation outcomes into bandit feedback using final managed-population hypervolume gain, diversity gain, rank smoothing, and invalid-code penalties. The profiler records budget curves, AUBC, bandit logs, and final population artifacts for analysis.

At a broader level, the project shows that MPaGE-style heuristic design can be strengthened by making LLM-call allocation adaptive. MPaGE already provides a productive foundation through PFG selection, semantic clustering, and prompt-based variation. The contribution of \texttt{mpage\_bmab} is to place these mechanisms under a controller that learns which operators and semantic regions deserve additional budget during the run.

\section{Main Contributions}

The first contribution is the formulation of LLM-based heuristic design as a budget-constrained sequential decision problem. Rather than treating the budget merely as a stopping condition, the project makes budget consumption part of the algorithmic state and evaluates progress over the entire budget trajectory through AUBC.

The second contribution is a two-level bandit scheduler for MPaGE-style heuristic generation. The operator bandit learns long-term preferences between mutation and crossover, while the cluster bandit adapts to the short-lived semantic clusters constructed in each generation. This structure is well matched to the project setting: operators are stable across the run, whereas cluster identities are local and non-stationary.

The third contribution is the integration of non-stationarity handling through Page--Hinkley drift detection. Because a semantic cluster can become less useful after repeated exploitation, resetting stale cluster statistics is a practical mechanism for avoiding persistent overconfidence in outdated reward estimates.

The fourth contribution is the final-HV-oriented reward and population-handling design. The finalized method rewards the actual generated child, aligns the quality signal with the managed final-population hypervolume, prevents the last affordable call from being spent only on clustering, incorporates front-aware cluster priors, weights parents by PFG membership, and flushes pending valid offspring into the final population before result writing.

The fifth contribution is an experimental and analysis pipeline that records not only terminal results but also the process by which budget is consumed. Artifacts such as \code{budget\_curve.json}, \code{aubc.json}, \code{bandit\_log.json}, validity summaries, and the generated thesis figures under \sourcefile{mpage\_bmab/thesis/figures/} support analysis of both final quality and budget efficiency.

\section{Main Findings}

The current result set provides strong practical evidence that BMAB-style scheduling improves budget efficiency over the original MPaGE workflow. Across the four-setup matrix, the Final-HV reward method wins 15 of 16 AUBC task--budget cells against MPaGE-orig, while Dense reward and Hybrid reward each win all 16 AUBC cells against MPaGE-orig. The largest qualitative differences appear in tasks where the baseline budget curve rises later, particularly Tri-TSP, Bi-CVRP, and Bi-KP.

The evidence for final hypervolume is also favorable to the Final-HV reward method, but it remains more nuanced than AUBC. This setup has the best mean final-HV rank, six best final-HV cells, and twelve top-two final-HV cells. It also wins 13 of 16 final-HV cells against MPaGE-orig. Nevertheless, Dense reward, Hybrid reward, and occasionally MPaGE-orig remain competitive in some terminal cells. Therefore, the correct conclusion is not uniform terminal domination, but improved budget trajectory with generally strong terminal quality.

The reward-mode comparison clarifies an important trade-off. Hybrid reward obtains the best valid-yield profile and a slightly higher normalized AUBC score than Final-HV reward, whereas Final-HV reward has the strongest final-HV rank. Dense reward remains competitive but is weaker than Final-HV reward in final-HV score. These results support the final-HV-oriented reward as the safest official default while showing that dense and hybrid reward signals remain useful for early progress and valid candidate production.

The task-level analysis gives a more nuanced view. On Bi-TSP, several methods reach similar final-HV values, so the BMAB advantage is expressed mainly through better budget trajectory. On Tri-TSP, final-HV values are also close, but BMAB-style methods reach high outer HV earlier. On Bi-CVRP, the AUBC gap against MPaGE-orig is especially large, suggesting that adaptive scheduling is valuable when route feasibility and structured local moves make generation more fragile. On Bi-KP, hybrid reward is particularly competitive, which suggests that dense local feedback may be useful for binary item-selection representations.

The most accurate empirical conclusion is therefore careful rather than absolute. The proposed method should not be described as uniformly dominating every setup on every metric. It should be described as consistently improving budget-integrated performance while maintaining strong terminal quality. This interpretation matches the research objective more closely than a final-HV-only summary would.

\section{Limitations}

The thesis is limited by the available experimental evidence. Each task--budget cell contains five seeds, which is too small for strong statistical conclusions under a two-sided Wilcoxon signed-rank test. The reported trends are broad and consistent, but they should be interpreted as empirical evidence rather than definitive statistical proof.

The current complete result matrix includes the MPaGE baseline, the official Final-HV reward method, and two reward-mode variants. However, it does not include complete matrices for all component ablations. The codebase implements additional variants, including \texttt{no\_budget\_anneal}, \texttt{no\_ph}, \texttt{no\_diversity}, \texttt{op\_only}, and \texttt{cluster\_only}. Without complete result matrices for these variants, the thesis cannot isolate the causal contribution of every scheduler component.

The repository does not provide hardware specifications, GPU usage, memory usage, API latency, token-level cost, or monetary cost for the reported runs. No information was found in the project source code or documentation for these measurements. Consequently, the current budget model counts LLM calls rather than modeling heterogeneous token costs or real-time latency.

Finally, the reported HV and AUBC values are outer heuristic-population metrics. They are meaningful for comparing budget-aware heuristic-design processes under the project harness, but they are not a direct reproduction of all inner-task solution-quality tables from the original MPaGE paper.

The thesis is also limited by the absence of qualitative code analysis. The saved generated samples make it possible to inspect recurring heuristic motifs, invalid-generation patterns, or semantic differences between clusters, but the current study focuses on quantitative outer-population metrics. Such qualitative analysis would be useful for explaining why particular reward modes or task families behave differently.

\section{Future Work}

The most immediate future direction is to run a complete component ablation matrix. The reward-mode comparison is now available for \texttt{full}, \texttt{dense\_reward}, and \texttt{hybrid\_reward}, but the effects of Page--Hinkley drift detection, diversity reward, operator-only scheduling, cluster-only scheduling, and budget annealing still require complete controlled sweeps.

A second direction is to increase the number of seeds. More seeds would improve the power of statistical tests and make it easier to distinguish robust effects from noise caused by stochastic LLM generation, random initial archives, and generated heuristic randomness.

A third direction is to extend the budget model from uniform call cost to token- or latency-aware cost. In practical LLM deployments, prompts and completions can differ substantially in length, and different models can have different price and latency profiles. A token-aware budgeted bandit would better reflect deployment constraints.

A fourth direction is to log and analyze invalid-generation failures more precisely. Separating syntax errors, missing function definitions, runtime exceptions, infeasible outputs, and timeouts would support targeted prompt engineering and repair strategies. Such diagnostics could also clarify whether validity improvements or reward allocation contribute more to AUBC gains.

Finally, the search controller could be extended with higher-level planning mechanisms, such as prompt repair, memory over successful heuristic motifs, or structured exploration over operator sequences. These extensions should be evaluated carefully because any additional LLM calls must justify their cost under the same AUBC-oriented budget criterion.

A sixth direction is adaptive reward scheduling. The current results show that dense reward is useful for early progress and that final-HV reward is strongest for terminal quality. This suggests a natural schedule in which the controller uses more dense or hybrid feedback early and gradually shifts toward final-HV feedback as the remaining budget decreases.

A seventh direction is qualitative heuristic analysis. Since the repository stores generated samples, future work can cluster or manually inspect the generated code to identify recurring local-search structures. This could answer questions that pure HV metrics cannot answer, such as whether successful heuristics share common move operators, whether invalid code arises from specific prompt patterns, or whether crossover actually combines useful motifs from two parents.

An eighth direction is token-aware and cost-aware budgeting. The current budget model counts LLM calls uniformly, but real model usage depends on prompt length, completion length, latency, and model pricing. A future cost-aware controller could treat mutation, crossover, and clustering as actions with variable monetary or token costs. This would make the budgeted-bandit formulation closer to real deployment conditions.

\section{Final Remarks}

\texttt{mpage\_bmab} demonstrates that budget allocation is a substantive algorithmic issue in LLM-driven heuristic design. By adding a two-level bandit scheduler, hard budget accounting, drift detection, and budget-curve profiling to MPaGE, the project turns a fixed-budget search loop into an adaptive resource-allocation process. The current experiments indicate that this design substantially improves budget-integrated performance, especially on more complex benchmark families. Although further ablations and larger seed counts are required for stronger causal and statistical claims, the project provides a coherent foundation for studying Generative AI heuristic design under realistic cost constraints.
